### 2013-03-23 eusflat2013/introduction.tex
Almost the last changes! Text is too long though :s

Information systems (IS) have always tried to model parts of reality. To achieve this modelling, such systems contain data representing properties of real-world objects or concepts. As time is an essential aspect of many real-world objects or concepts, information systems often contain data representing temporal notions which describe such temporal properties. Such temporal notions usually take the form of either instants~\cite{Dyreson1994},~\cite{Jensen1998}, which can informally be seen as infinitely short `moments' or `points' in time, or time intervals~\cite{Dyreson1994},~\cite{Jensen1998}.

Data are often produced by humans, but human-made data are prone to imperfections: some data may be vague, imprecise, incomplete, contradictory or uncertain. Data representing temporal notions may contain such imprecisions too. The work presented in this paper is specifically concerned with time intervals (and as a special case: instants) subject to uncertainty.

@@

Of course, when temporal information is represented in an IS, querying this IS may have a temporal aspect too. Usually, querying an IS containing data representing temporal notions is conceptually done by specifying one or more time indications and requesting information that is in a specific relationship with these indications, where the semantics of these relationships are specifically temporal. Thus, some existing proposals have considered groups of basic relationships between time indications used to construct and express specific temporal relationships. Notably, Allen~\cite{Allen1983} presented a reasoning framework containing all semantically usefull basic temporal relationships between time intervals (and as a special case instants). The resulting relationships are shown in figure \ref{fig:allen-relationships}. These Allen relationships are used in the presented work.

\begin{figure}[h]
   \centering
   \includegraphics[width=0.8\columnwidth]{graphs/allen.eps}
   \caption{The Allen relationships between two crisp intervals $I$ and $J$.  }
   \label{fig:allen-relationships}
 \end{figure}

To be able to query IS containing data representing time indications subject to uncertainty, a framework is necessary, able to represent uncertainties in time indications in a semantically sound way, without (much) information loss and able to temporally reason with such time indications in a semantically sound and usefull way. Although more proposals for such frameworks exist, the work presented in this paper focusses on just two: the \emph{ill-known constraint \emph{(IKC)} framework}~\cite{Pons2011} and the \emph{triangular model \emph{(TM)} framework}~\cite{DeTre2012}.

>>>>

Information systems (IS) have always tried to model parts of reality. To achieve this modelling, such systems contain data representing properties of real-world objects or concepts~\cite{JoseEnriquePons2012},~\cite{Billiet2012}. As time is an essential aspect of many real-world objects or concepts, information systems often contain data representing temporal notions which describe such temporal properties~\cite{Bolour1982},~\cite{VanderCruyssen1997}. Such temporal notions usually take the form of either instants~\cite{Jensen1998}, which can informally be seen as infinitely short `moments' or `points' in time, or time intervals~\cite{Jensen1998}.

Data are often produced by humans, but human-made data are prone to imperfections: some data may be vague, imprecise, incomplete, contradictory or uncertain. Data representing temporal notions may contain such imprecisions too~\cite{Devos1994},~\cite{Dubois2003},~\cite{JoseEnriquePons2012},~\cite{Billiet2012}. The work presented in this paper is specifically concerned with time intervals (and as a special case: instants) subject to uncertainty.

@@

Of course, when temporal information is represented in an IS, querying this IS may have a temporal aspect too. Usually, querying an IS containing data representing temporal notions is conceptually done by specifying one or more time indications and requesting information that is in a specific relationship with these indications, where the semantics of these relationships are specifically temporal~\cite{Billiet2012},~\cite{JoseEnriquePons2012},~\cite{Pons2012a}. Thus, some existing proposals have considered groups of basic relationships between time indications used to construct and express specific temporal relationships~\cite{Medina1994},~\cite{Schockaert2008}. Notably, Allen~\cite{Allen1983} presented a reasoning framework containing all semantically usefull basic temporal relationships between time intervals (and as a special case instants). The resulting relationships are shown in figure \ref{fig:allen-relationships}. These Allen relationships are used in the presented work.

\begin{figure}[h]
   \centering
   \includegraphics[width=0.9\columnwidth]{graphs/allen.eps}
   \caption{The Allen relationships between two crisp intervals $I$ and $J$.  }
   \label{fig:allen-relationships}
 \end{figure}

To be able to query IS containing data representing time indications subject to uncertainty, a framework is necessary, able to represent uncertainties in time indications in a semantically sound way, without (much) information loss and able to temporally reason with such time indications in a semantically sound and usefull way~\cite{Dubois1983},~\cite{Dubois2003}. Although more proposals for such frameworks exist, the work presented in this paper focusses on just two: the \emph{ill-known constraint \emph{(IKC)} framework}~\cite{Pons2011} and the \emph{triangular model \emph{(TM)} framework}~\cite{DeTre2012}.

### 2013-03-23 eusflat2013/tm.tex
Almost the last changes! Text is too long though :s

\subsection{\label{subsec:tm-preliminaries}Preliminary Concepts}
In the triangular model framework, the approaches to interval representation and Allen relationship evaluation are closely linked with the approach to interval visualization.

Usually, different time intervals are visualized as different parallel line segments in the same image plane. However, as this linear approach introduces a few issues, time intervals are visualized as points in the image plane in the TM framework.

To achieve this, a horizontal line segment visualizing the used (part of the) time axis is drawn in the image plane. Then, a triangle is drawn, using this line segment as a side of which the angles with the other two sides have sizes $\alpha$ and $-\alpha$ respectively. The area contained in this triangle is called the \emph{interval space} and will contain all visualizations of intervals. Now, the procedure used to visualize a crisp time interval $\left[s, e\right]$ in the interval space is the following:

@@

A similar point visualizing an interval is generally called an \emph{interval point}. In this work, as it is usually done, the size of $\alpha$ is chosen to be $45^{\circ}$. An example of such construction is shown in figure \ref{fig:tm-const-ex}.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\columnwidth]{graphs/TM_model_several.eps}
	\caption{The visualization of several crisp time intervals using the TM framework.}
	\label{fig:tm-const-ex}
\end{figure}

\subsection{\label{subsec:tm-interval}Representation of Time Intervals Subject to Uncertainty}
The TM framework allows uncertainty in time intervals by supporting uncertain time intervals:

@@

Consider an ordered set of instants $T$ and an UTI $J$ in $T$, defined by the pair $(\pi_s, \pi_e)$. Given an instant $x \in T$, $\pi_s(x)$ now expresses the possibility that $x$ is the starting instant of $J$ and $\pi_e(x)$ expresses the possibility that $x$ is the ending instant of $J$. Thus, $J$ intends to indicate just one crisp time interval, but due to a (partial) lack of knowledge, it is (partially) unknown exactly which time interval this is. Because $\pi_s$ and $\pi_e$ are required to be convex, the possibility $\pi_J(I)$ that a given crisp time interval $I = \left[s_i, e_i\right]$ is the exact time interval intended by $J$ can now be uniquely determined by:

@@

Now consider an ordered set of instants $T$ and a visualization of this in an image plane, using the TM framework. An UTI $J$ in $T$ defined by a pair $(\pi_s, \pi_e)$ is now visualized as an area in the image plane. This area is constructed as follows: for every crisp time interval $K$ for which the possibility $\pi_J(K)$ of $K$ being the time interval represented by $J$ is higher than zero, the interval point is drawn and is given a color. This color is part of a linear single-color greyscale and its saturation is linearly related to the degree of possibility that $K$ is the time interval intended by $J$. Thus, every drawn interval point in this visualization corresponds to an interval with a higher-than-zero possibility of being the time interval intended by $J$. Together, these drawn interval points form the visualization of $J$. An example of such construction is shown in figure \ref{fig:tm-ill-const-ex}. In this figure, a darker color of interval point indicates a higher and a lighter color a lower possibility that the corresponding interval is the intended interval.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\columnwidth]{graphs/TM_model_ill_known.eps}
	\caption{The visualization of one UTI defined by a pair $(\pi_s, \pi_e)$ using the TM framework.}
	\label{fig:tm-ill-const-ex}
\end{figure}

\subsection{\label{subsec:tm-evaluation}Evaluation of Allen Relationships}
At the base of the framework's approach to evaluating Allen relationships is the concept of uncertain relational zones:

@@

Now, given an UTI $J$ and its visualization in an interval space, the procedure to visualize its URZ is the following:

@@

In this procedure, $\alpha$ is the same angle size as in the procedure for the visualization of crisp time intervals. The results of this procedure for an example UTI $J$ (as wel as the visualization of $J$ itself) are shown in figure \ref{fig:tm-urz}.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\columnwidth]{graphs/TM_model_URZ.eps}
	\caption{The visualization of the URZ for a given UTI $J$, using the TM framework.}
	\label{fig:tm-urz}
\end{figure}

Given an UTI $J$ and its URZ, every URZ is assigned a symbol and a name and corresponds to a set of Allen relationships. These symbols and names are used to uniquely identify each of $J$'s URZ. The set of Allen relationships corresponding to an URZ contains every Allen relationship in which any interval visualized by one of the URZ's interval points could be with the interval intended by $J$, depending on exactly which time interval $J$ actually intends to represent. In figure \ref{fig:tm-urz}, the URZ's symbols are noted inside the areas visualizing their respective URZ's. In table \ref{tab:urz}, every different row corresponds to a different URZ. In the first two columns, the symbol and name of the URZ are given, the last column contains the set of corresponding Allen relationships. For these, the following notations are used:

@@

\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|}
\hline
Symbol & Name & Relationships \\
\hline
B    & Before & B \\
O    & Overlaps & O \\
C    & Contains & C \\
D    & During & D \\
OB   & Overlapped by & OB \\
A    & After & A \\
PM   & Possibly meets & M, B, O \\
PS   & Possibly starts & O, S, D \\
PFB  & Possibly finished by & O, FB, C \\
\multirow{2}{*}
{PE}   & Possibly equal  & E, FB, SB, F,\\
       &                 & S, C, D,O, OB \\
PSB  & Possibly started by & C, OB, SB \\
PF   & Possibly finishes & OB, F, D \\
PMB  & Possibly met by & OB, MB A \\
UB   & Uncertain beginning & B, M, O, S, D \\
UE   & Uncertain ending & D, F, OB, MB, A \\
\hline
\end{tabular}
\caption{The fifteen possible URZ for a given UTI.}
\label{tab:urz}
\end{table}

The TM framework now allows evaluating the Allen relationships a given crisp time interval $I = \left[s, e\right]$ could be in with a given UTI $J$, depending on the exact interval intended by $J$. This is done as follows. First, $J$ and its URZ are visualized in an interval space. Next, $I$'s interval point, its construction half-lines $L_s$ and $L_e$ and the visualization of its starting and ending instants on the visualization of the time are drawn in the same interval space. Then, evaluation is a matter of position (in the following, the possibility that $I$ is the interval intended by $J$ is denoted $\pi_J(I)$ and the possibility that $I$ is in Allen relationship $AR$ with $J$ is denoted  $\Pos(I\ AR\ J)$):

@@

Here, the intervals $K$ are crisp time intervals and the expression `$K \in p_i$' is a notation to express that $K$ is visualized by an interval point contained in area part $p_i$.

>>>>

\subsection{\label{subsec:tm-preliminaries}Preliminary Concepts}
In the triangular model framework, the approaches to interval representation and Allen relationship evaluation are closely related to the approach to interval visualization. Usually, different time intervals are visualized as different parallel line segments in the same image plane. However, as this linear approach introduces a few issues, time intervals are visualized as points in the image plane in the TM framework~\cite{Weghe2007},~\cite{DeTre2012}. To achieve this, a horizontal line segment visualizing the used (part of the) time axis is drawn in the image plane. Then, a triangle is drawn, using this line segment as a side of which the angles with the other two sides have sizes $\alpha$ and $-\alpha$ respectively. The area contained in this triangle is called the \emph{interval space} and will contain all visualizations of intervals. Now, the procedure used to visualize a CTI $\left[s, e\right]$ in the interval space is the following~\cite{Weghe2007},~\cite{DeTre2012}:

@@

A similar point visualizing an interval is generally called an \emph{interval point}~\cite{Weghe2007},~\cite{DeTre2012}. In this work, as it is usually done, the size of $\alpha$ is chosen to be $45^{\circ}$. An example of such construction is shown in figure \ref{fig:tm-const-ex}.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\columnwidth]{graphs/TM_model_several.eps}
	\caption{The visualization of several CTI using the TM framework.}
	\label{fig:tm-const-ex}
\end{figure}

\subsection{\label{subsec:tm-interval}Representation of Time Intervals Subject to Uncertainty}
The TM framework allows uncertainty in time intervals by supporting uncertain time intervals~\cite{DeTre2012}:

@@

Consider an ordered set of instants $T$ and an UTI $J$ in $T$, defined by the pair $(\pi_s, \pi_e)$. Given an instant $x \in T$, $\pi_s(x)$ now expresses the possibility that $x$ is the starting instant of $J$ and $\pi_e(x)$ expresses the possibility that $x$ is the ending instant of $J$. Thus, $J$ intends to indicate just one CTI, but due to a (partial) lack of knowledge, it is (partially) unknown exactly which time interval this is. Because $\pi_s$ and $\pi_e$ are required to be convex, the possibility $\pi_J(I)$ that a given CTI $I = \left[s_i, e_i\right]$ is the exact time interval intended by $J$ can now be uniquely determined by~\cite{DeTre2012}:

@@

Now consider an ordered set of instants $T$ and a visualization of this in an image plane, using the TM framework. An UTI $J$ in $T$ defined by a pair $(\pi_s, \pi_e)$ is now visualized as an area in the image plane. This area is constructed as follows: for every CTI $K$ for which the possibility $\pi_J(K)$ of $K$ being the time interval represented by $J$ is higher than zero, the interval point is drawn and is given a color. This color is part of a linear single-color greyscale and its saturation is linearly related to the degree of possibility that $K$ is the time interval intended by $J$. Thus, every drawn interval point in this visualization corresponds to an interval with a higher-than-zero possibility of being the time interval intended by $J$. Together, these drawn interval points form the visualization of $J$~\cite{DeTre2012}. An example of such construction is shown in figure \ref{fig:tm-ill-const-ex}. In this figure, a darker color of interval point indicates a higher and a lighter color a lower possibility that the corresponding interval is the intended interval.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\columnwidth]{graphs/TM_model_ill_known.eps}
	\caption{The visualization of one UTI defined by a pair $(\pi_s, \pi_e)$ using the TM framework.}
	\label{fig:tm-ill-const-ex}
\end{figure}

\subsection{\label{subsec:tm-evaluation}Evaluation of Allen Relationships}
At the base of the framework's approach to evaluating Allen relationships is the concept of uncertain relational zones~\cite{DeTre2012}:

@@

Now, given an UTI $J$ and its visualization in an interval space, the procedure to visualize its URZ is the following~\cite{DeTre2012}:

@@

In this procedure, $\alpha$ is the same angle size as in the procedure for the visualization of CTI. The results of this procedure for an example UTI $J$ (as wel as the visualization of $J$ itself) are shown in figure \ref{fig:tm-urz}.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\columnwidth]{graphs/TM_model_URZ.eps}
	\caption{The visualization of the URZ for a given UTI $J$, using the TM framework.}
	\label{fig:tm-urz}
\end{figure}

Given an UTI $J$ and its URZ, every URZ is assigned a symbol and a name and corresponds to a set of Allen relationships~\cite{DeTre2012}. These symbols and names are used to uniquely identify each of $J$'s URZ. The set of Allen relationships corresponding to an URZ contains every Allen relationship in which any interval visualized by one of the URZ's interval points could be with the interval intended by $J$, depending on exactly which time interval $J$ actually intends to represent. In figure \ref{fig:tm-urz}, the URZ's symbols are noted inside the areas visualizing their respective URZ's. In table \ref{tab:urz}, every different row corresponds to a different URZ. In the first two columns, the symbol and name of the URZ are given, the last column contains the set of corresponding Allen relationships. For these, the following notations are used:

@@

\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|}
\hline
Symbol & Name & Relationships \\
\hline
B    & Before & B \\
O    & Overlaps & O \\
C    & Contains & C \\
D    & During & D \\
OB   & Overlapped By & OB \\
A    & After & A \\
PM   & Possibly Meets & M, B, O \\
PS   & Possibly Starts & O, S, D \\
PFB  & Possibly Finished By & O, FB, C \\
\multirow{2}{*}
{PE}   & Possibly Equal  & E, FB, SB, F,\\
       &                 & S, C, D,O, OB \\
PSB  & Possibly Started By & C, OB, SB \\
PF   & Possibly Finishes & OB, F, D \\
PMB  & Possibly Met By & OB, MB A \\
UB   & Uncertain Beginning & B, M, O, S, D \\
UE   & Uncertain Ending & D, F, OB, MB, A \\
\hline
\end{tabular}
\caption{The fifteen possible URZ for a given UTI.}
\label{tab:urz}
\end{table}

The TM framework now allows evaluating the Allen relationships a given CTI $I = \left[s, e\right]$ could be in with a given UTI $J$, depending on the exact interval intended by $J$. This is done as follows. First, $J$ and its URZ are visualized in an interval space. Next, $I$'s interval point, its construction half-lines $L_s$ and $L_e$ and the visualization of its starting and ending instants on the visualization of the time are drawn in the same interval space. Then, evaluation is a matter of position (in the following, the possibility that $I$ is the interval intended by $J$ is denoted $\pi_J(I)$ and the possibility that $I$ is in Allen relationship $AR$ with $J$ is denoted  $\Pos(I\ AR\ J)$)~\cite{DeTre2012}:

@@

Here, the intervals $K$ are CTI and the expression `$K \in p_i$' is a notation to express that $K$ is visualized by an interval point contained in area part $p_i$.
